Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-22 demystify.tex
updates

\chapter*{Preface}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students} (and also domestic ones), especially those in  smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

@@

\begin{commentbox}[Vu:]
  If your grading system is not US standard or you are from a good school unknown outside of your country, you can ask your reference writers to explain about that in their letters.  For example, "Bach Khoa" are the top universities in Vietnam for STEM studies but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that.
\end{commentbox}

@@

\paragraph{School} Graduating from top universities \emph{that we recognize} helps. For example, if you are an international student and your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English.

\textbf{Advice}: In general, just treat the interview as an informal chat, but be prepared to talk about your research experiences and interests (so do have \emph{elevator pitch} about your experiences and research).
A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). Some prof. can ask you to solve a problem, e.g., a coding problem, so you might want to ask if you need to prepare for that.
During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

@@

\epigraph{"Oh... and how is education supposed to make me feel smarter? Besides, every time I learn something new, it pushes some old stuff out of my brain. Remember when I took that home wine-making course and I forgot how to drive?"}{\textsc{The Simpsons}}

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra "activity" fee or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

@@

\begin{commentbox}[Hung:]
  I always encourage the students I admitted to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a professor’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their "level of excitement" being in the group.
\end{commentbox}

@@

\subsubsection*{Additional Resources}
\begin{itemize}
  \item \href{https://www.cs.columbia.edu/wp-content/uploads/2019/03/Get-Advisor.pdf}{The Definitive "what do I ask/look for" in a PhD Advisor Guide}
\end{itemize}
\subsection{Types of Faculty: Who can serve as a PhD adviser?}\label{sec:faculty-types}

@@

Faculty often receive "cold" e-mails from prospective students. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that get our attention?

@@

\chapter{Miscs and FAQs}
\epigraph{"I want to share something with you – the three little sentences that will get you through life; number 1: Cover for me, number 2: Oh, good idea, Boss, and number 3: It was like that when I got here."}{\textsc{The Simpsons}}

\section{Why did I get rejected?}\label{sec:why-rejected}
Many possible reasons!  Other than not having a strong profile (e.g., research potentials, GPA, LoRs, SOP, poor interviews), here are some that you might not expect:

@@

@@

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are red flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in short time). If you think you have red flags, you should address them in your SOP.
In general, these things are rare, but they do happen and cause concern to the adcom.

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google) (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

In his \href{https://matt.might.net/articles/how-to-apply-and-get-in-to-graduate-school-in-science-mathematics-engineering-or-computer-science/}{post}, Matt Might was initially unsure about an application. However, upon learning that the applicant had led a 100km hike in the Himalayas, he decided to accept the applicant.  This is a good example of being "stand out", and I would also advocate for that student as this shows they have the persistence and determination required for research.

@@

\section{Can I apply to CS PhD if my undergrad was not in CS or related areas?}\label{sec:non-stem}

Yes, \emph{as long as} you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out as mentioned in \autoref{sec:improve-your-chance}.

One main concern for non-CS or non-STEM students is if you have the sufficient technical background obtained through core CS courses.  You need to show that you have such knowledge through your coursework, projects, or research.
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your SOP.  If you have done a project that requires knowledge of OS or have a professional certification (e.g., A+) through work, you can talk about it.  If you have done research that requires knowledge of Discrete Maths, you can talk about it.  You can also ask your LoR writers to talk about your technical background.
In summary, in your application, convince us that you have the background to do CS PhD research.

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have these essential knowledge, e.g., through the mentioned ways. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a "bridge" courses to help students catch up.  For example, GMU has four bridge courses (Data Structures and Algorithms, Computer Systems, Discrete Math, and Programming Foundation) that incoming students can take to catch up on their CS knowledge.

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student can get MS degree ``along the way'' to PhD, e.g., after finishing course work in the first 2 years.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution.

@@

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you (who already have an MS) start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of "exams", e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word "ABD" (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you might not have these exams or only have to do a few of them.

@@

\emph{Faculty position after PhD}: In other countries, PhD graduates interested academia typically apply for additional research appointments, i.e., postdocs, and then consider faculty position. In the US, PhD graduates often apply directly for faculty position (postdoc for US graduates is no longer a popular option as it was before).  %This is also why US PhD programs are longer (remember, 5--7 years).

@@

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their names if they are OK with it as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{"I did my postdoc with Jeff Foster at Univ. of Maryland"}.

>>>>

\chapter*{Preface}
Having been involved in PhD admissions for many years, I've realized that many \textbf{international} (and domestic) students, especially those in  smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

@@

\begin{commentbox}[Vu:]
  If your grading system is not US standard or you are from a good school unknown outside of your country, you can ask your reference writers to explain about that in their letters.  For example, ``Bach Khoa'' are the top universities in Vietnam for STEM studies but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that.
\end{commentbox}

@@

\paragraph{School} Graduating from top universities \emph{that we recognize} helps. For example, if you are an international student and your school is well-known, then it is \emph{``top foreign''}, which is a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
You might be asked to talk about your research experiences and interests, and to read a paper and discuss it. In some rare cases you might also be asked to solve a problem, e.g., a coding problem. 
The interview is also a good opportunity for you to ask questions about the program and the university.

\textbf{Advice}: Treat the interview as an informal chat. Have elevator pitches about your research experiences and interests. You might also want have some 5-minute presentation about your research. If a professor asks you to read a paper, do it and be prepared to discuss it. You can also ask if you need to prepare for a coding problem. Finally, prepare some thoughtful questions to ask, e.g., about the program, the university, or the professor's research.

@@

\epigraph{``Oh... and how is education supposed to make me feel smarter? Besides, every time I learn something new, it pushes some old stuff out of my brain. Remember when I took that home wine-making course and I forgot how to drive?''}{\textsc{The Simpsons}}

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra activity fee or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

@@

\begin{commentbox}[Hung:]
  I always encourage the students I admitted to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a professor’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their ``level of excitement'' being in the group.
\end{commentbox}

@@

\subsubsection*{Additional Resources}
\begin{itemize}
  \item \href{https://www.cs.columbia.edu/wp-content/uploads/2019/03/Get-Advisor.pdf}{The Definitive ``what do I ask/look for'' in a PhD Advisor Guide}
\end{itemize}
\subsection{Types of Faculty: Who can serve as a PhD adviser?}\label{sec:faculty-types}

@@

Faculty often receive ``cold'' e-mails from prospective students. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that get our attention?

@@

\chapter{Miscs and FAQs}
\epigraph{``I want to share something with you – the three little sentences that will get you through life; number 1: Cover for me, number 2: Oh, good idea, Boss, and number 3: It was like that when I got here.''}{\textsc{The Simpsons}}

\section{Why did I get rejected?}\label{sec:why-rejected}
Many possible reasons!  Other than not having a strong profile (e.g., research potentials, GPA, LoRs, SOP, poor interviews), here are some that you might not expect.

@@

\paragraph{You aim too high}  You might have applied to schools that are too high for your profile. While being ambitious is good, you should also be realistic and apply to a range of schools, including ``safety'' ones.  I have seen many students who have excellent profiles but only applied to top 10 or even top 20 schools and got rejected from all of them.  Before applying, you should talk to your professors and ask them to give you an honest opinion on where you should apply

@@

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are \red{red} flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in short time). If you think you have red flags, you should address them in your SOP.
In general, these things are rare, but they do happen and cause concern to the adcom.

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for ``URM minority in CS in the US'' on Google) (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

In his \href{https://matt.might.net/articles/how-to-apply-and-get-in-to-graduate-school-in-science-mathematics-engineering-or-computer-science/}{post}, Matt Might was initially unsure about an application. However, upon learning that the applicant had led a 100km hike in the Himalayas, he decided to accept the applicant.  This is a good example of being \emph{stand out}, and I would also advocate for that student as this shows they have the persistence and determination required for research.

@@

\section{My undergrad was not in CS or related areas}\label{sec:non-stem}

You still can apply to PhD in CS \emph{as long as} you can demonstrate you are ready for it through your background, research experiences, LoRs, statements, etc. You might be even able to leverage this to make your profile stand out as mentioned in \autoref{sec:improve-your-chance}.

A main question adcom has for a non-CS or non-STEM student is if you have the sufficient technical background obtained through core CS courses.  So you need to show that you have such knowledge through your coursework, projects, or research.
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your SOP.  If you have done a project that requires knowledge of OS or have a professional certification (e.g., A+) through work, you can talk about it.  If you have done research that requires knowledge of Discrete Maths, you can talk about it.  You can also ask your LoR writers to talk about your technical background.
In summary, in your application, convince us that you have the background to do CS PhD research.

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have these essential knowledge, e.g., through the mentioned ways. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, GMU has four bridge courses (Data Structures and Algorithms, Computer Systems, Discrete Math, and Programming Foundation) that incoming students can take to catch up on their CS knowledge.

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student can get MS degree ``along the way'' to PhD, e.g., after finishing course work in the first 2 years.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

@@

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you (who already have an MS) start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of exams, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you might not have these exams or only have to do a few of them.

@@

\emph{Faculty position after PhD}: In other countries, PhD graduates interested academia typically apply for additional research appointments, i.e., postdocs, and then consider faculty position. In the US, PhD graduates often apply directly for faculty position (postdoc for US graduates is no longer a popular option as it was before).

@@

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their names if they are OK with it as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{``I did my postdoc with Jeff Foster at Univ. of Maryland''}.
